---
title: "Two-Compartment IV Infusion"
format:
  html:
    code-fold: false  
    code-summary: "Code"
    code-tools: true
    number-sections: true
    toc-depth: 5
editor_options: 
  chunk_output_type: console
---

```{css, echo = FALSE}
.scrolling_500 {
  max-height: 500px;
  overflow-y: auto;
}

.scrolling_700 {
  max-height: 700px;
  overflow-y: auto;
}

.cell-output {
  background-color: WhiteSmoke;
  border-radius: 4px;
}
```

Here, we write down the PK model for a two-compartment model with IV infusion 
and linear elimination. We also write down the statistical model for multiple 
error models. The Stan models are free to download using the corresponding 
download buttons. 

# PK Model

The data-generating model is 
$$C_{ij} = f(\mathbf{\theta}_i, t_{ij}) + error$$
where $error$ can be

*  additive: $C_{ij} = f(\mathbf{\theta}_i, t_{ij}) + \epsilon_{a_{ij}}$
*  proportional: $C_{ij} = f(\mathbf{\theta}_i, t_{ij})*\left( 1 + \epsilon_{p_{ij}} \right)$
*  proportional-plus-additive: $C_{ij} = f(\mathbf{\theta}_i, t_{ij})*\left( 1 + \epsilon_{p_{ij}} \right) + \epsilon_{a_{ij}}$
*  lognormal: $C_{ij} = f(\mathbf{\theta}_i, t_{ij})*e^{\epsilon_{ij}}$

where $\mathbf{\theta}_i = \left[CL_i, \, V_{c_i}, \, Q_i, \,V_{p_i}\right]^\top$ 
is a vector containing the individual parameters for individual $i$[^poppk-4], 
$f(\cdot, \cdot)$ is the two-compartment model mean function seen [here](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/PFIM3.2.2/PFIM_PKPD_library.pdf?root=pkpd), and $\epsilon_p, \epsilon_a,$ and $\epsilon$ are normally distributed. The
corresponding system of ODEs is:

[^poppk-4]: These parameters are the macro-parameterization of this model - 
clearance, central compartment volume, inter-compartmental clearance, and 
peripheral compartment volume, respectively. I have chosen to use $V_c$ and 
$V_p$ for **c**entral compartment volume and **p**eripheral compartment volume 
rather than $V_1$ and $V_2$ or $V_2$ and $V_3$ to reduce any confusion about the
numbering (since $V_2$ is the central compartment in ADVAN4 but the peripheral 
compartment in ADVAN3 in NONMEM)

$$\begin{align}
\frac{dA_c}{dt} &= -\left(\frac{CL}{V_c} + \frac{Q}{V_c}\right)A_C + \frac{Q}{V_p}A_p + rate_{in} \notag \\
\frac{dA_p}{dt} &= \frac{Q}{V_c}A_c - \frac{Q}{V_p}A_p \\
\end{align}$$ {#eq-2cmt-macro-genode-no-depot}

and then $C = \frac{A_c}{V_c}$[^poppk-5].

[^poppk-5]: For this model, the matrix-exponential is faster than the 
analytical solution with Torsten, so I've written the Stan models to use the 
linear ODE solution. Therefore, since this is purely IV, I've written the ODEs
with two compartments. If you want to use the analytical solution, you would 
write the ODEs as
$$\begin{align}
\frac{dA_d}{dt} &= -K_a*A_d \notag \\
\frac{dA_c}{dt} &= K_a*A_d - \left(\frac{CL}{V_c} + \frac{Q}{V_c}\right)A_C + \frac{Q}{V_p}A_p + rate_{in} \notag \\
\frac{dA_p}{dt} &= \frac{Q}{V_c}A_c - \frac{Q}{V_p}A_p \\
\end{align}$$ {#eq-2cmt-macro-genode-with-depot}
You'll notice that this is written as if there is first-order absorption. This 
is to emphasize that the central compartment is the second compartment in 
Torsten's setup, and you would have to make sure $CMT = 2$ for the dosing and 
observation records. Since there is no absorption, we just leave it out of our 
model and set it to 0 when using Torsten's analytical solver.

# Statistical Model

I'll model the *correlation matrix* $R_{\Omega}$ and *standard deviations* 
($\omega_p$) of the random effects rather than the covariance matrix $\Omega$ 
that is the more traditional method. 

\begin{align}
\Omega &=
\begin{pmatrix}
\omega^2_{CL} & \omega_{CL, V_c} & \omega_{CL, Q} & \omega_{CL, V_p} \\
\omega_{CL, V_c} & \omega^2_{V_c} & \omega_{V_c, Q} & \omega_{V_c, V_p} \\
\omega_{CL, Q} & \omega_{V_c, Q} & \omega^2_{Q} & \omega_{Q, V_p} \\
\omega_{CL, V_p} & \omega_{V_c, V_p} & \omega_{Q, V_p} & \omega^2_{V_p} \\
\end{pmatrix} \\
&= \begin{pmatrix}
\omega_{CL} & 0 & 0 & 0 \\
0 & \omega_{V_c} & 0 & 0 \\
0 & 0 & \omega_{Q} & 0 \\
0 & 0 & 0 & \omega_{V_p} \\
\end{pmatrix} \mathbf{R_{\Omega}}
\begin{pmatrix}
\omega_{CL} & 0 & 0 & 0 \\
0 & \omega_{V_c} & 0 & 0 \\
0 & 0 & \omega_{Q} & 0 \\
0 & 0 & 0 & \omega_{V_p} \\
\end{pmatrix} \\
&= \begin{pmatrix}
\omega_{CL} & 0 & 0 & 0 \\
0 & \omega_{V_c} & 0 & 0 \\
0 & 0 & \omega_{Q} & 0 \\
0 & 0 & 0 & \omega_{V_p} \\
\end{pmatrix}
\begin{pmatrix}
1 & \rho_{CL, V_c} & \rho_{CL, Q} & \rho_{CL, V_p} \\
\rho_{CL, V_c} & 1 & \rho_{V_c, Q} & \rho_{V_c, V_p} \\
\rho_{CL, Q} & \rho_{V_c, Q} & 1 & \rho_{Q, V_p} \\
\rho_{CL, V_p} & \rho_{V_c, V_p} & \rho_{Q, V_p} & 1 \\
\end{pmatrix}
\begin{pmatrix}
\omega_{CL} & 0 & 0 & 0 \\
0 & \omega_{V_c} & 0 & 0 \\
0 & 0 & \omega_{Q} & 0 \\
0 & 0 & 0 & \omega_{V_p} \\
\end{pmatrix}
\end{align} 

When the random effects are modeled this way, we get all of the same information
as if they were modeled with a full covariance matrix, *i.e.*, we can transform
the standard deviations and correlation matrix into the covariance matrix if we 
want. In addition, standard deviations and correlations are more interpretable
than variances and covariances. It's also much easier to set a prior 
distribution on the standard deviations and the correlation matrix with an 
[LKJ](https://en.wikipedia.org/wiki/Lewandowski-Kurowicka-Joe_distribution) 
distribution than it is to set a prior on a covariance matrix with an 
[inverse-Wishart](https://en.wikipedia.org/wiki/Inverse-Wishart_distribution)
distribution.


The full statistical model is 


::: panel-tabset
## Proportional Error Model

\begin{align}
C_{ij} \mid \mathbf{TV}, \; \mathbf{\eta}_i, \; \mathbf{\Omega}, \; \mathbf{\Sigma}
&\sim Normal\left( f(\mathbf{\theta}_i, t_{ij}), \; \sigma_{ij} \right)
I(C_{ij} > 0) \notag \\
\mathbf{\eta}_i \; | \; \Omega &\sim Normal\left(
\begin{pmatrix}
0 \\ 0 \\ 0 \\ 0 \\
\end{pmatrix}
, \; \Omega\right) \notag \\
TVCL &\sim Lognormal\left(log\left(location_{TVCL}\right), scale_{TVCL}\right) \notag \\
TVVC &\sim Lognormal\left(log\left(location_{TVVC}\right), scale_{TVVC}\right) \notag \\
TVQ &\sim Lognormal\left(log\left(location_{TVQ}\right), scale_{TVQ}\right) \notag \\
TVVP &\sim Lognormal\left(log\left(location_{TVVP}\right), scale_{TVVP}\right) \notag \\
\omega_{CL} &\sim Half-Normal(0, scale_{\omega_{CL}}) \notag \\
\omega_{V_c} &\sim Half-Normal(0, scale_{\omega_{V_c}}) \notag \\
\omega_{Q} &\sim Half-Normal(0, scale_{\omega_{Q}}) \notag \\
\omega_{V_p} &\sim Half-Normal(0, scale_{\omega_{V_p}}) \notag \\
R_{\Omega} &\sim LKJ(df_{R_{\Omega}}) \notag \\
\sigma_p &\sim Half-Normal(0, scale_{\sigma_p}) \\
\end{align}

where 
\begin{align}
\mathbf{TV} &=
\begin{pmatrix}
TVCL \\ TVVC \\ TVQ \\ TVVP \\
\end{pmatrix} \\
\mathbf{\theta}_i &=
\begin{pmatrix}
CL_i \\ V_{c_i} \\ Q_i \\ V_{p_i} \\
\end{pmatrix} =
\begin{pmatrix}
TVCL \times e^{\eta_{CL_i}} \\ TVVC \times e^{\eta_{V_{c_i}}} \\ TVQ \times e^{\eta_{Q_i}} \\ TVVP \times e^{\eta_{V_{p_i}}} \\
\end{pmatrix} \\
\mathbf{\eta}_i &=
\begin{pmatrix}
\eta_{CL_i} \\ \eta_{V_{c_i}} \\ \eta_{Q_i} \\ \eta_{V_{p_i}} \\
\end{pmatrix} \\
\sigma_{ij} &= f(\mathbf{\theta}_i, t_{ij})\sigma^2_p  \\
\end{align}

Note: The indicator for $C_{ij} | \ldots$ indicates that we are truncating the 
distribution of the observed concentrations to be greater than 0.


## Proportional-plus-Additive Error Model

\begin{align}
C_{ij} \mid \mathbf{TV}, \; \mathbf{\eta}_i, \; \mathbf{\Omega}, \; \mathbf{\Sigma}
&\sim Normal\left( f(\mathbf{\theta}_i, t_{ij}), \; \sigma_{ij} \right)
I(C_{ij} > 0) \notag \\
\mathbf{\eta}_i \; | \; \Omega &\sim Normal\left(
\begin{pmatrix}
0 \\ 0 \\ 0 \\ 0 \\
\end{pmatrix}
, \; \Omega\right) \notag \\
TVCL &\sim Lognormal\left(log\left(location_{TVCL}\right), scale_{TVCL}\right) \notag \\
TVVC &\sim Lognormal\left(log\left(location_{TVVC}\right), scale_{TVVC}\right) \notag \\
TVQ &\sim Lognormal\left(log\left(location_{TVQ}\right), scale_{TVQ}\right) \notag \\
TVVP &\sim Lognormal\left(log\left(location_{TVVP}\right), scale_{TVVP}\right) \notag \\
\omega_{CL} &\sim Half-Normal(0, scale_{\omega_{CL}}) \notag \\
\omega_{V_c} &\sim Half-Normal(0, scale_{\omega_{V_c}}) \notag \\
\omega_{Q} &\sim Half-Normal(0, scale_{\omega_{Q}}) \notag \\
\omega_{V_p} &\sim Half-Normal(0, scale_{\omega_{V_p}}) \notag \\
R_{\Omega} &\sim LKJ(df_{R_{\Omega}}) \notag \\
\sigma_p &\sim Half-Normal(0, scale_{\sigma_p}) \\
\sigma_a &\sim Half-Normal(0, scale_{\sigma_a}) \\
R_{\Sigma} &\sim LKJ(df_{R_{\Sigma}}) \notag \\
\end{align}

where 
\begin{align}
\mathbf{TV} &=
\begin{pmatrix}
TVCL \\ TVVC \\ TVQ \\ TVVP \\
\end{pmatrix} \\
\mathbf{\theta}_i &=
\begin{pmatrix}
CL_i \\ V_{c_i} \\ Q_i \\ V_{p_i} \\
\end{pmatrix} =
\begin{pmatrix}
TVCL \times e^{\eta_{CL_i}} \\ TVVC \times e^{\eta_{V_{c_i}}} \\ TVQ \times e^{\eta_{Q_i}} \\ TVVP \times e^{\eta_{V_{p_i}}} \\
\end{pmatrix} \\
\mathbf{\eta}_i &=
\begin{pmatrix}
\eta_{CL_i} \\ \eta_{V_{c_i}} \\ \eta_{Q_i} \\ \eta_{V_{p_i}} \\
\end{pmatrix} \\
\mathbf{\Sigma} &=
\begin{pmatrix}
\sigma^2_{p} & \rho_{p,a}\sigma_{p}\sigma_{a} \\
\rho_{p,a}\sigma_{p}\sigma_{a} & \sigma^2_{z} \\
\end{pmatrix} \\
\sigma_{ij} &= \sqrt{f(\mathbf{\theta}_i, t_{ij})^2\sigma^2_p + \sigma^2_a + 2f(\mathbf{\theta}_i, t_{ij})\rho_{p,a}\sigma_{p}\sigma_{a}} \\
\end{align}

Note: The indicator for $C_{ij} | \ldots$ indicates that we are truncating the 
distribution of the observed concentrations to be greater than 0.

## Lognormal Error

\begin{align}
C_{ij} \mid \mathbf{TV}, \; \mathbf{\eta}_i, \; \mathbf{\Omega}, \; \mathbf{\Sigma}
&\sim Lognormal\left( log\left(f(\mathbf{\theta}_i, t_{ij})\right), \; \sigma \right)
\notag \\
\mathbf{\eta}_i \; | \; \Omega &\sim Normal\left(
\begin{pmatrix}
0 \\ 0 \\ 0 \\ 0 \\
\end{pmatrix}
, \; \Omega\right) \notag \\
TVCL &\sim Lognormal\left(log\left(location_{TVCL}\right), scale_{TVCL}\right) \notag \\
TVVC &\sim Lognormal\left(log\left(location_{TVVC}\right), scale_{TVVC}\right) \notag \\
TVQ &\sim Lognormal\left(log\left(location_{TVQ}\right), scale_{TVQ}\right) \notag \\
TVVP &\sim Lognormal\left(log\left(location_{TVVP}\right), scale_{TVVP}\right) \notag \\
\omega_{CL} &\sim Half-Normal(0, scale_{\omega_{CL}}) \notag \\
\omega_{V_c} &\sim Half-Normal(0, scale_{\omega_{V_c}}) \notag \\
\omega_{Q} &\sim Half-Normal(0, scale_{\omega_{Q}}) \notag \\
\omega_{V_p} &\sim Half-Normal(0, scale_{\omega_{V_p}}) \notag \\
R_{\Omega} &\sim LKJ(df_{R_{\Omega}}) \notag \\
\sigma &\sim Half-Normal(0, scale_{\sigma}) \\
\end{align}

where 
\begin{align}
\mathbf{TV} &=
\begin{pmatrix}
TVCL \\ TVVC \\ TVQ \\ TVVP \\
\end{pmatrix} \\
\mathbf{\theta}_i &=
\begin{pmatrix}
CL_i \\ V_{c_i} \\ Q_i \\ V_{p_i} \\
\end{pmatrix} =
\begin{pmatrix}
TVCL \times e^{\eta_{CL_i}} \\ TVVC \times e^{\eta_{V_{c_i}}} \\ TVQ \times e^{\eta_{Q_i}} \\ TVVP \times e^{\eta_{V_{p_i}}} \\
\end{pmatrix} \\
\mathbf{\eta}_i &=
\begin{pmatrix}
\eta_{CL_i} \\ \eta_{V_{c_i}} \\ \eta_{Q_i} \\ \eta_{V_{p_i}} \\
\end{pmatrix} \\
\end{align}


:::


# Stan Code

## Proportional Error

```{r}
#| echo: false

downloadthis::download_file(
  path = here::here("models/stan/iv_2cmt_prop.stan"),
  output_name = "iv_2cmt_prop",
  button_label = "Download 2-cmt IV with proportional error",
  button_type = "default",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

```

::::{.scrolling_700}

:::{.cell-output}

```{r}
#| echo: false
#| output: true

readLines(here::here("models/stan/iv_2cmt_prop.stan")) |>
  writeLines()

```

:::

::::

## Proportional-Plus-Additive Error

```{r}
#| echo: false

downloadthis::download_file(
  path = here::here("models/stan/iv_2cmt_ppa.stan"),
  output_name = "iv_2cmt_ppa",
  button_label = "Download 2-cmt IV with proportional-plus-additive error",
  button_type = "default",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

```

::::{.scrolling_700}

:::{.cell-output}

```{r}
#| echo: false
#| output: true

readLines(here::here("models/stan/iv_2cmt_ppa.stan")) |>
  writeLines()

```

:::

::::

## Lognormal Error

```{r}
#| echo: false

downloadthis::download_file(
  path = here::here("models/stan/iv_2cmt_exp.stan"),
  output_name = "iv_2cmt_exp",
  button_label = "Download 2-cmt IV with lognormal error",
  button_type = "default",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

```

::::{.scrolling_700}

:::{.cell-output}

```{r}
#| echo: false
#| output: true

readLines(here::here("models/stan/iv_2cmt_exp.stan")) |>
  writeLines()

```

:::

::::